Liveness-Based Garbage Collection for Lazy Languages

### 2015-12-01 paper/paper.tex

Memory  usage graphs  for select  benchmarks are  shown
Figure~\ref{fig:memory-usage},  Column   (a).   In some cases, the
number of  garbage collections  tend to be  very large,
hence  we  show only  a  slice of the original graph  with 
representative    behavior  (Column (b)).  In all the programs we can see
that  the   curve  corresponding  to  LGC   (\LGCLine\ line)
dips   below the RGC curve 
(\RGCLine\ line) during garbage collection.  The graphs also include the
curve    for   reachable    cells   (\ReachLine)  and live cells
(\UseLine). These were  obtained
 by forcing  RGC to  run at   very  high
frequency. The curve for live cells were obtained by recording heap
access times and post processing the data 
at the end of the program.

As demonstrated by the gap between the \RGCLine\ and the \UseLine, a large number of cells 
which are unused by the program are still copied during RGC. LGC does a much better job of closing this gap 
but still falls short of the precision achieved by LGC in case of eager languages
~\cite{asati14lgc}. 
A case for concern is the huge difference in the number of cells touched by LGC vs RGC for some programs.
This is due to heap locations being  traversed multiple times during LGC. \warning{We have implemented heuristics to 
overcome common cases and they show promising results}.

@@

\warning{Although we  provide an  implementation for  a liveness
based garbage  collection scheme for a lazy  language, we
do not  provide a formal  proof of its  correctness.  A
formal proof of correctness would describe how closures
are handled  during garbage collection and  liveness is
propagated inside closures.

  Another  interesting exercise  is to  narrow the  gap
  between  the   actual  liveness  and   the  perceived
  liveness  of our  garbage collector. Using
  strictness  analysis  to  eagerly  evaluate  closures
  might release more dead cells to be garbage
  collected.  Since our  garbage collector  can handle
  closures   (suspended  evaluations),   another  major
  challenge  is to  extend it  to support  higher order
  programs.

  Orthogonally, we  plan to  improve the  efficiency of
  the liveness based garbage collector using heuristics
  such   as  limiting   the  depth   of  DFA,   merging
  nearly-equivalent    states    and    using    better
  representation    and    algorithms   for    automata
  manipulation.  We   also  need  to   investigate  the
  interaction   of  liveness   with  other   collection
  schemes,   such  as   incremental  and   generational
  collection.  In summary, we  need to investigate ways
  to make liveness  based garbage collection attractive
  for practical collectors.}

>>>>

Memory  usage graphs  for select  benchmarks are  shown
Figure~\ref{fig:memory-usage},  Column   (a).   In some cases, the
number of  garbage collections  tend to be  very large,
hence  we  show only  a  slice of the original graph  with 
representative    behavior  (Column (b)).  In all the programs we can see
that  the   curve  corresponding  to  LGC   (\LGCLine\ line)
dips   below the RGC curve 
(\RGCLine\ line) during garbage collection.  The graphs also include the
curve    for   reachable    cells   (\ReachLine)  and live cells
(\UseLine). These were  obtained
 by forcing  RGC to  run at   very  high
frequency. The curve for live cells were obtained by recording heap
access times and post processing the data 
at the end of the program. Note that the size of an LGC cell is 1.16
times the size of a RGC cell.

As demonstrated by the gap between the \RGCLine\ and the \UseLine, a large number of cells 
which are unused by the program are still copied during RGC. LGC does a much better job of closing this gap 
but still falls short of the precision achieved by LGC in case of eager languages
~\cite{asati14lgc}. 
A major source of inefficiency in LGC is multiple traversals of
already copied heap cells. Since LGC does not mark the heap cells
after the first visit, the same cells can be repeatedly visited with
different liveness states. We have mitigated this problem by
implementing a heuristic which tags \CONS\ cells with the set of
states already visited.

@@

The graphs show that even in lazy languages there are large amount of
dead cells that can be collected early. In spite of collecting
closures there still exists a gap between actual and perceived
livenesses.
Language features that could be introduced are higher-order functions
and recursive lets. We believe that functions can be used to model
recursive lets and this should be a simple extension to our liveness
analysis.
For higher-order we intend to use firstification techniques~\cite{Mitchell:2009}.

  Orthogonally, we  plan to  improve the  efficiency of
  the liveness based garbage collector using heuristics
  such   as  limiting   the  depth   of  DFA,   merging
  nearly-equivalent    states    and    using    better
  representation(for example BDDs~\cite{Bryant86})    and    algorithms   for    automata
  manipulation.  We   also  need  to   investigate  the
  interaction   of  liveness   with  other   collection
  schemes,   such  as   incremental  and   generational
  collection.  It might be interesting to use a mixed mode garbage collection
  scheme which allows the costs of LGC to be amortized over several runs
  of RGC. In summary, we  need to investigate ways
  to make liveness  based garbage collection attractive
  for practical collectors.
